Light inextensible strings (thread) under tension in the Schwarzschild geometry

### 2021-01-27 physics_educator_rebuttal.txt

>  
> 8. A radius of \sqrt{27M} (page 4 line 9) What radius is this?

Derivation added as a result of point 7 above should clarify this
issue.

>  
> 9. Although photon paths provide a definition of
> straightness... (discussion section line 2) In GR, worldlines of
> photon are null geodesics. The paper does not distinguish geodesics of
> spatial sections and geodesics of spacetime.
>

I would argue that emphasising the concept of "spatial sections" might
be somewhat counterproductive, even though I admittedly present a
diagram of Flamm's paraboloid [which is obligatory here, I would say].
A spatial section S is a 3D subset of a 4D spacetime in which we
require the squared interval between any two distinct events of S to
be spacelike.  It is easy to give an example of a spatial section in
flat Minkowski space: with standard coordinates (t,x,y,z) we consider
events (t0,x,y,z) for some fixed t0.  Such constructions are somewhat
more difficult in Schwarzschild space but we might use Schwarzschild
coordinates (t,r,theta,phi) and then specify S2=(t0,r,theta,phi).  The
Schwarzschild metric then shows that S2 is a spatial section, provided
that r>2M, because the Schwarzschild metric (equation 1 in the
manuscript) is diagonal, and the nonvanishing components are strictly
positive.

When considering static strings it is easier and more convenient to
use standard Schwarzschild coordinates (t,r,theta,phi) and then simply
ignore the 't' coordinate as it plays no part in calculating proper
length of a string (being careful to specify that we are outside the
event horizon).  If the string is static, we may imagine two static
observers tied to the string; in calculating the amount of proper
distance between the two observers (string length) we simply ignore
the 't' coordinate and integrate the spatial component of the
Schwarzschild metric.  This process is distinct from the concept of
spatial section in which we condition on t=t0.

The world lines of freely-falling objects follow non-null geodesics
and photons follow null geodesics.  Light inextensible string follows
neither of these paths.  I think the revised manuscript makes this
clear

### 2021-01-27 physics_educator_rebuttal.txt

Rebuttal follows.  Reviewers' comments preceded by a chevron (">") and
my response included inline.

Overall response: the comments are perfectly reasonable and I have
addressed them individually.  The manuscript is much stengthened as a
result and I commend it to you.

@@

> Reviewers' comments:
>  
> Reviewer 1: This paper studies geodesics on the equatorial plane of
> the spatial sections (constant time slices) of the Schwarzschild
> spacetime. The geodesic equation in this case reduces to a second
> order non-linear ODE, given by Eq. (6) in the paper, and it is solved
> numerically. A few geodesics are plotted in Fig. 1, 2, 3, and 5.
>  
> The author's idea is to use the path of light taut strings to
> visualize "straight lines in the vicinity of a black hole". The paper
> claims that light taut strings will follow paths that minimize proper
> length, so the paths are geodesics in spatial sections of the
> Schwarzschild spacetime in the second paragraph in the introduction
> section and also in the discussion section. However, this major claim
> was made without any justification. Justification for this major claim
> needs to be given.
>

On the one hand, this is a perfectly reasonable request: the claim is
indeed a major one, and one that had no explicit justification in the
manuscript.  But I could argue that it requires no justification: one
could say that the claim is true by definition.  Didactically, one
might conceptualise a string as being comprised of many tiny links,
each of which contributes some small length to the string, and each of
which is sufficiently small for Euclidean geometry to describe its
length.  Each link is subject to a well-defined local tension with
respect to its neighbours.

@@

We may then *count* the links and use this as a proxy for proper
length, demonstrating the consistency of the concept of proper length;
and of course small links oblige one to use infinitesimal calculus.

However, much of the above reasoning is necessarily tautological; it
is perfectly reasonable to *define* an "inextensible light string
under tension" as one that adopts a minimal [or at least extremal]
proper-length path between its end points.  I have included a short
discussion in the manuscript.

@@

> The paper has imprecise usage of GR terminology. A few statements are
> unclear. Examples include:
>  
> 1.  "straight line" in the vicinity of black
> hole... (Introduction/paragraph 1) I believe it refers to straight
> line in spatial sections of black hole.
>

The original sentence was:

:: Here I show how considering light inextensible 
:: string can be used to interpret the concept of "straight line" in the
:: vicinity of a black hole.

It might be rephrased: "...interpret the concept of "straight line" in
spatial sections of a black hole", but I believe this form of words
would be potentially misleading.  My thinking was that the concept of
"straight line" ---in the abstract sense of how straightness itself is
defined---is to be re-interpreted in the manuscript as the path taken
by a light inextensible string.

@@

That is an extremely good observation.  The word "large" is
inappropriate and misleading here, I think.  Asymptotic expansions can
be confusing and what I meant to say was that the ratio of proper
string length to Euclidean string length (that is, length as estimated
by the Euclidean formula (1+r)-1 where 'r' is circumferential radius)
approaches infinity.  But the word "large" is not right here, and I
have rephrased accordingly.

I have also realised that this equation and the preceding one for
small w can be made much more comprehensible by giving an example, and
I have included two examples using solar-mass black holes.

>  
> 3. with radial strings being arbitrarily longer than flat space would
> allow... (below Eq. 4) what is the meaning of this statement?
>

Euclidean geometry predicts that l=2Mw; the intent of that sentence
was to point out that we Euclidean predictions are arbirtarily
incorrect as we move away from the black hole.  I have rephrased and
included a numerical example which should appeal to students of GR.

@@

>  
> 7. Radius of curvature (page 4 line 7) Since the equatorial plane is
> now curved, what radius of curvature is the author talking about?
> Clarification is needed.

On re-reading, the original manuscript was a little cryptic.  I have
added a detailed derivation of this result that should clarify the
issue.  In essence, a light beam and a tangential string will diverge
quadratically and this leads to the radius of curvature quoted.  The
manuscript is much clearer than before.

>  
> 8. A radius of \sqrt{27M} (page 4 line 9) What radius is this?

Derivation added as a result of point 7 above should clarify this
issue.  This is the curvature of the string that a static observer,
who is tied to the string, would measure if he considered a beam of
light to be "straight".

@@

>  
> 9. Although photon paths provide a definition of
> straightness... (discussion section line 2) In GR, worldlines of
> photon are null geodesics. The paper does not distinguish geodesics of
> spatial sections and geodesics of spacetime.
>

I would argue that emphasising the concept of "spatial sections" might
be somewhat counterproductive, even though I admittedly present a
diagram of Flamm's paraboloid [which is obligatory here, I would say].
A spatial section S is a 3D subset of a 4D spacetime in which we
require the squared interval between any two distinct events of S to
be spacelike.  It is easy to give an example of a spatial section in
flat Minkowski space: with standard coordinates (t,x,y,z) we consider
events (t0,x,y,z) for some fixed t0.  Such constructions are somewhat
more difficult in Schwarzschild space but we might use Schwarzschild
coordinates (t,r,theta,phi) and then specify S2=(t0,r,theta,phi).  The
Schwarzschild metric then shows that S2 is a spatial section, provided
that r>2M, because the Schwarzschild metric (equation 1 in the
manuscript) is diagonal, and the nonvanishing components are strictly
positive.

When considering static strings it is easier and more convenient to
use standard Schwarzschild coordinates (t,r,theta,phi) and then simply
ignore the 't' coordinate as it plays no part in calculating proper
length of a string (being careful to specify that we are outside the
event horizon).  If the string is static, we may imagine two static
observers tied to the string; in calculating the amount of proper
distance between the two observers (string length) we simply ignore
the 't' coordinate and integrate the spatial component of the
Schwarzschild metric.  This process is distinct from the concept of
spatial section in which we condition on t=t0.

The world lines of freely-falling objects follow non-null geodesics
and photons follow null geodesics.  Light inextensible string follows
neither of these paths.  I hope that the revised manuscript makes this
clear, and the various diagrams of strings and null geodesics should
emphasise the difference between the two.

### 2021-02-24 tpe_edreview_rebuttal.txt

Review of "Light inextensible strings (thread) under tension in the
Schwarzschild geometry" submitted to The Physics Educator

I want to begin by congratulating the author on a very clever
exercise.  If the route of a non-stretchable string under tension
would take the path of least distance, asking about the behavior of a
tight string in a Schwarzschild geometry, and comparing that path to
that of a light ray, offers an interesting question.  One might
initially suppose the two paths would be the same, or approximately
so.  The author rightly observes that General Relativity is a "topic
of long-standing fascination for physics undergraduates." True, but I
am unsure whether this particular exercise will offer more
understanding or more confusion for the intended undergraduates,
because the distinction between a stretched string and a null metrics
may be blurred (see point 11 below).

I would point out that at least one calculation presented in the paper
can found in existing introductory General Relativity textbooks.
Specifically, the meaning of the r-coordinate, and the difference
between Delta r and the distance between two concentric shells of
different r-coordinates (eq. 2) is thoroughly discussed in Taylor &
Wheeler's book Exploring Black Holes: Introduction to General
Relativity (Addison-Wesley-Longman, 2000), Ch. 2.  While the
calculation of Eq. (2) should be included in this paper for
completeness and done before going on to other calculations, Taylor &
Wheeler's black holes book is a standard pedagogical reference for
such calculations, and should be included among the references.
Taylor & Wheeler's book is very likely the intended undergraduate
student's first encounter with General Relativity.

The paper raises a few questions in this reviewer's mind, and in
elsewhere it offers opportunities for suggestions.

1.  Question.  First paragraph in "Introduction": The third sentence
"Here I interpret straight line as the path adopted by a stationary,
light inextensible string under tension passing between two points."
Is this to be understood as identical to the definition of "geodesic"?
If that is the assumption, can it be justified, because when one
thinks of geodesic in General Relativity, one has "free-fall" in mind,
and the stretched string, being "stationary," would not be in
free-fall in the global coordinate system.

2.  Suggestion about the line that begins on p. 1 and continues to
p. 2, about "lengthscales $L\leq T/(\mu g)$: Saying it this way may
present a sticking point for the undergraduate audience towards whom
this paper in intended.  The logic behind this inequality would show
greater clarity if one says something like "the weight of the string
does not exceed the tension, so that $L\mu g<T$".  From this the
stated inequality about lengthscales follows easily.

3.  Suggestion.  Stating "$c^4/4GM$ for the strength of gravity at the
event horizon" may leave an undergraduate audience baffled.  By
starting from where they are, with $g=GM/r^2$, then setting r equal to
the Schwarzschild radius, the stated result follows immediately,
without mystery.

4.  Question. Lines 3-4, page 2.  Can the author explain what is meant
by "the dominant energy condition"?

5.  Suggestion.  2nd and 3rd line below eq. (2): the statement "A
circular string at constant r will, trivially, have proper length
$2\pi r$."  Stating the relation this way muddies the logic and the
distinction between flat and Schwarzschild space.  The r-coordinate of
a spherical shell centered on a point mass M is defined operationally
by measuring the circumference of the shell and dividing the
circumference's value by $2\pi$.  That must be understood before the
calculations of Eq. (2) can make sense.  As the sentence is presently
structured it may obscure the distinction between the r coordinate and
radial lengths because it invites the undergraduate reader to
interpret the "triviality" as a statement about plane Euclidean
geometry instead of the Schwarzschild geometry's definition of the
r-coordinate.

6.  Question.  There may be a typographical error in Eq. (2).
Shouldn't the coefficient of the logarithm be 2M instead of 1/2?  As
written the equation is not dimensionally consistent, since r and M
are measured in the same units.  Compare to Taylor & Wheeler's
Exploring Black Holes, p. 2-28, Eq. (27).

7.  Suggestion: It appears that the $L$ and the $l$ that appear
between Eq. (1) and Eq. (3) denote the same quantity.  Why use
different symbols?

8.  Suggestion: In the line above Eq. (3), if $w >> 0$ then one does
not need absolute value bars in $|w|<< 1$.

9.  Question.  2nd line below Eq. (4), why are radial strings
"arbitrarily" longer than flat space would imply?  It appears the
difference is a matter of calculation (or measurement), not arbitrary.
Please explain this choice of words.

10. Suggestion.  Line below Eq. (5), I think the word "dashes" should
be "primes".  This is a dash: --- and this is a prime: ', at least in
usage with which I am familiar.

11.  The paragraph below Eq. (6) is puzzling.  Here the path of the
tight string is being compared to an "equatorial null geodesic," which
I take to mean the path of a light beam moving along the equator.  If
we are considering a light beam, then shouldn't a term with $t$ in it
appear somewhere?  The puzzlement continues given the statement that
"light, being sensitive to the time component of the metric" whereas
the equations discussed in this paragraph say nothing (that I can see)
about the time component of the metric.  The proper time interval
would be zero, but the coordinate time interval $dt$ would, I think,
be nonzero.  What am I missing?

12.  In the passage below Eq. (6), I am having trouble making the
connection between $r''(\phi)=r-3M + 2(r')^2/r$ and the claim that the
solution is $r=r_0\sec(\phi-\phi_0)$.  What trig identity am I
missing?

13.  Can the author explain more fully what is meant by a "tangential
null geodesic" (several lines after Eq. 6)?  That terminology makes it
sound like the r-coordinate does not change, whence $r'$ would be
zero, but that appears to not be the case.  In this section, if $r'$
is small I see where the approximation $r''=r-2M$ comes from, but I am
having trouble seeing where $r''-r-3M$ arises for the alleged
tangential null geodesics.  What am I not seeing?

I will have to take the author's word for the numerical results.

This is an interesting paper.  I congratulate the author for thinking
of such an intriguing application of the Schwarzschild metric.
However, in this reviewer's opinion, before the paper can be well
fitted to its announced audience ("an undergraduate homework problem")
some points need to be clarified.

### 2021-02-24 tpe_edreview_rebuttal.txt

Herewith my rebuttal to the review of "Light inextensible strings..."
The referee's comments are preceded by a chevron (">") and my response
is inline.  I have tinkered with linebreaks here and there in the
interests of clarity, and have replaced some mathematical notation in
the original review with latex equivalents.

Short story: I agree with all the comments made and have modified the
manuscript accordingly; changes appear in the manuscript using colour
coding for ease of reference.  The document is stronger, more
cohesive, more accurate, and hopefully more instructive; and I commend
it to you.

> Review of "Light inextensible strings (thread) under tension in the
> Schwarzschild geometry" submitted to The Physics Educator
> 
> I want to begin by congratulating the author on a very clever
> exercise.  If the route of a non-stretchable string under tension
> would take the path of least distance, asking about the behavior of a
> tight string in a Schwarzschild geometry, and comparing that path to
> that of a light ray, offers an interesting question.  One might
> initially suppose the two paths would be the same, or approximately
> so.  The author rightly observes that General Relativity is a "topic
> of long-standing fascination for physics undergraduates." True, but I
> am unsure whether this particular exercise will offer more
> understanding or more confusion for the intended undergraduates,
> because the distinction between a stretched string and a null metrics
> may be blurred (see point 11 below).
> 
> I would point out that at least one calculation presented in the paper
> can found in existing introductory General Relativity textbooks.
> Specifically, the meaning of the r-coordinate, and the difference
> between Delta r and the distance between two concentric shells of
> different r-coordinates (eq. 2) is thoroughly discussed in Taylor &
> Wheeler's book Exploring Black Holes: Introduction to General
> Relativity (Addison-Wesley-Longman, 2000), Ch. 2.  While the
> calculation of Eq. (2) should be included in this paper for
> completeness and done before going on to other calculations, Taylor &
> Wheeler's black holes book is a standard pedagogical reference for
> such calculations, and should be included among the references.
> Taylor & Wheeler's book is very likely the intended undergraduate
> student's first encounter with General Relativity.
> 
> The paper raises a few questions in this reviewer's mind, and in
> elsewhere it offers opportunities for suggestions.
> 
> 1.  Question.  First paragraph in "Introduction": The third sentence
> "Here I interpret straight line as the path adopted by a stationary,
> light inextensible string under tension passing between two points."
> Is this to be understood as identical to the definition of "geodesic"?
> If that is the assumption, can it be justified, because when one
> thinks of geodesic in General Relativity, one has "free-fall" in mind,
> and the stretched string, being "stationary," would not be in
> free-fall in the global coordinate system.
> 
> 2.  Suggestion about the line that begins on p. 1 and continues to
> p. 2, about "lengthscales $L\leq T/(\mu g)$: Saying it this way may
> present a sticking point for the undergraduate audience towards whom
> this paper in intended.  The logic behind this inequality would show
> greater clarity if one says something like "the weight of the string
> does not exceed the tension, so that $L\mu g<T$".  From this the
> stated inequality about lengthscales follows easily.
> 
> 3.  Suggestion.  Stating "$c^4/4GM$ for the strength of gravity at the
> event horizon" may leave an undergraduate audience baffled.  By
> starting from where they are, with $g=GM/r^2$, then setting r equal to
> the Schwarzschild radius, the stated result follows immediately,
> without mystery.
> 
> 4.  Question. Lines 3-4, page 2.  Can the author explain what is meant
> by "the dominant energy condition"?
> 
> 5.  Suggestion.  2nd and 3rd line below eq. (2): the statement "A
> circular string at constant r will, trivially, have proper length
> $2\pi r$."  Stating the relation this way muddies the logic and the
> distinction between flat and Schwarzschild space.  The r-coordinate of
> a spherical shell centered on a point mass M is defined operationally
> by measuring the circumference of the shell and dividing the
> circumference's value by $2\pi$.  That must be understood before the
> calculations of Eq. (2) can make sense.  As the sentence is presently
> structured it may obscure the distinction between the r coordinate and
> radial lengths because it invites the undergraduate reader to
> interpret the "triviality" as a statement about plane Euclidean
> geometry instead of the Schwarzschild geometry's definition of the
> r-coordinate.
> 
> 6.  Question.  There may be a typographical error in Eq. (2).
> Shouldn't the coefficient of the logarithm be 2M instead of 1/2?  As
> written the equation is not dimensionally consistent, since r and M
> are measured in the same units.  Compare to Taylor & Wheeler's
> Exploring Black Holes, p. 2-28, Eq. (27).
> 
> 7.  Suggestion: It appears that the $L$ and the $l$ that appear
> between Eq. (1) and Eq. (3) denote the same quantity.  Why use
> different symbols?
> 
> 8.  Suggestion: In the line above Eq. (3), if $w >> 0$ then one does
> not need absolute value bars in $|w|<< 1$.
> 
> 9.  Question.  2nd line below Eq. (4), why are radial strings
> "arbitrarily" longer than flat space would imply?  It appears the
> difference is a matter of calculation (or measurement), not arbitrary.
> Please explain this choice of words.
> 
> 10. Suggestion.  Line below Eq. (5), I think the word "dashes" should
> be "primes".  This is a dash: --- and this is a prime: ', at least in
> usage with which I am familiar.
> 
> 11.  The paragraph below Eq. (6) is puzzling.  Here the path of the
> tight string is being compared to an "equatorial null geodesic," which
> I take to mean the path of a light beam moving along the equator.  If
> we are considering a light beam, then shouldn't a term with $t$ in it
> appear somewhere?  The puzzlement continues given the statement that
> "light, being sensitive to the time component of the metric" whereas
> the equations discussed in this paragraph say nothing (that I can see)
> about the time component of the metric.  The proper time interval
> would be zero, but the coordinate time interval $dt$ would, I think,
> be nonzero.  What am I missing?
> 
> 12.  In the passage below Eq. (6), I am having trouble making the
> connection between $r''(\phi)=r-3M + 2(r')^2/r$ and the claim that the
> solution is $r=r_0\sec(\phi-\phi_0)$.  What trig identity am I
> missing?
> 
> 13.  Can the author explain more fully what is meant by a "tangential
> null geodesic" (several lines after Eq. 6)?  That terminology makes it
> sound like the r-coordinate does not change, whence $r'$ would be
> zero, but that appears to not be the case.  In this section, if $r'$
> is small I see where the approximation $r''=r-2M$ comes from, but I am
> having trouble seeing where $r''-r-3M$ arises for the alleged
> tangential null geodesics.  What am I not seeing?
> 
> I will have to take the author's word for the numerical results.
> 
> This is an interesting paper.  I congratulate the author for thinking
> of such an intriguing application of the Schwarzschild metric.
> However, in this reviewer's opinion, before the paper can be well
> fitted to its announced audience ("an undergraduate homework problem")
> some points need to be clarified.

### 2021-02-24 string_physics_educator_revision3.tex
start work on #22

\address{School of Engineering, Computing, and Mathematical Sciences, Auckland University of Technology, 55 Wellesley Street East\\
Auckland 1142, New Zealand
\email{hankin.robin@gmail.com}}

@@

Considering only stationary spacetimes, we define the proper length of
a string to be the result of integrating the spatial part of the
metric tensor along string's path between its endpoints.  A light
string's being under tension means that its path will be an extremum
of proper length: non-extremal paths between two fixed points cannot
be in stable equilibrium as the tension acts to shorten the path;
stable paths must therefore be of extremal proper-length.  In flat
space, the calculus of variations proves that the path is straight in
the sense of zero curvature.  In terrestrial gravity, standard
$25\times 10^{-6}\unit{kg}\cdot\unit{m}^{-1}$ cotton sewing thread at
its breaking strain of about $40\unit{N}$ would have a sag of~${\sim}
6\times 10^{-6}\unit{m}$ over a $1\unit{m}$ horizontal scale: barely
perceptible.  How does such a quotidian observation translate into the
vicinity of a black hole?  Would the ``warping of space'' result in
{\em visibly} curved thread?

@@

\noindent Here I consider the spatial component~$ds^2=
dr^2/\left(1-2M/r\right) + r^2d\phi^2$, confined to equatorial
points~($\theta=\pi/2$); we understand~$r\geqslant 2M$ throughout.  A
circular string at constant~$r$ will, trivially, have proper
length~$2\pi r$.

Considering first the radial component we may calculate the
proper length $L\left(r_1,r_2\right)$ of a string stretching
from~$r=r_1\geqslant 2M$ to~$r_2\geqslant r_1$ at constant $\theta,\phi$:

@@

\begin{equation}\label{radial_string_length}
  \int_{r=r_1}^{r_2}\frac{dr}{\sqrt{1-2M/r}}=
  \left.
  \sqrt{r(r-2M)} +\frac{1}{2}\log\left(
  \frac{\sqrt{r}+\sqrt{r-2M}}{\sqrt{r}-\sqrt{r-2M}}\right)
  \right|_{r_1}^{r_2}
  \end{equation}

Observe that integration may be performed with a lower limit
of~$r_1=2M$ with no difficulty, even though the radial component of the
Schwarzschild metric approaches infinity there.

Elementary methods show that the proper length $\ell$ of radial
strings has some non-Euclidean features.  Firstly, consider the proper
length from $r=2M$ to $r=2M(1+w)$ where $w>0, \left|w\right|\ll 1$.
Then

@@

(compare the flat space string length of $2Mw$).  This would indicate
 that, close to the Schwarzschild radius $r=2M$, more string is
 required to connect two points with slightly differing radial
 coordinates.  Taking a solar-mass black hole and $w=10^{-3}$ as an
 example, the proper length computed in flat space would be
 $10^{-2}\cdot 2M\simeq 3\,\mathrm{m}$---compare $\sim
 187\,\mathrm{m}$ from equation~\ref{ell}.  Passing to the case $w\gg
 1$ we have an asymptotic expansion that begins

@@

There does not appear to be a simple analytical solution to
equation~\ref{rdashdash}; but it is a straightforward second order
ODE, amenable to numerical solution and no harder than the equivalent
for null geodesics, a standard homework problem.  Here we consider
only strings exterior to the event horizon, $r>2M$, which have a
well-defined point of closest approach to the singularity.  Because
the system is invariant under~$\phi\longrightarrow-\phi$, the string
must be symmetrical about this point.  If $r'$ is small, then a string
follows $r''=r-2M$, while tangential null geodesics follow $r''=r-3M$.
Working to first order, and writing~$d$ for the difference in
Schwarzschild $r$ between string and null geodesic, we see that
$d''=M$.  Transforming coordinates to $(x,y) = (r\phi,r-r_0)$ and with
boundary conditions corresponding to a string and null geodesic
tangential at the $(x,y)$ origin, we see that
$d=\frac{M}{2}(x/r_0)^2$.  However, a static local observer tied to
the string would measure
$d_\mathrm{local}=d/\sqrt{1-2M/r_0}$---applying
equation~\ref{schwarzschild}---and obtain
$d_\mathrm{local}=\frac{M}{2}(x/r_0)^2\left(1-2M/r_0\right)^{-1/2}$
which would correspond to a radius of curvature
of~$r_0\cdot\frac{r_0}{M}\cdot\left(1-2M/r_0\right)^{-1/2}$.  Thus a
taut string would {\em look} curved to a static observer in
Schwarzschild coordinates, in the sense that a tangential light beam
and taut string would diverge quadratically.  At the photon sphere
$r_0=3M$, such a string would appear to have a radius of~$\sqrt{27}M$.
For a solar mass black hole this would be about $15\unit{km}$, a
readily understandable result.

>>>>

\address{School of Engineering, Computing, and Mathematical Sciences\\Auckland University of Technology\\55 Wellesley Street East\\
  Auckland CBD, Auckland 1010, New Zealand\\
\email{hankin.robin@gmail.com}}

@@

Considering only stationary spacetimes, we define the proper length of
a string to be the result of integrating the spatial part of the
metric tensor along string's path between its endpoints.  A light
string's being under tension means that its path will be an extremum
of proper length: non-extremal paths between two fixed points cannot
be in stable equilibrium as the tension acts to shorten the path;
stable paths must therefore be of extremal proper length.  In flat
space, the calculus of variations proves that the path is straight in
the sense of zero curvature.  In terrestrial gravity, standard
$25\times 10^{-6}\unit{kg}\cdot\unit{m}^{-1}$ cotton sewing thread at
its breaking strain of about $40\unit{N}$ would have a sag of~${\sim}
6\times 10^{-6}\unit{m}$ over a $1\unit{m}$ horizontal scale: barely
perceptible.  How does such a quotidian observation translate into the
vicinity of a black hole?  Would the ``warping of space'' result in
{\em visibly} curved thread?

@@

\noindent Here I consider the spatial component~$ds^2=
dr^2/\left(1-2M/r\right) + r^2d\phi^2$, confined to equatorial
points~($\theta=\pi/2$); we understand~$r\geqslant 2M$ throughout.  A
circular string at constant~$r$ will, trivially, have proper
length~$2\pi r$.  Considering first the radial component we may
calculate the proper length $L\left(r_1,r_2\right)$ of a string
stretching from~$r=r_1\geqslant 2M$ to~$r_2\geqslant r_1$ at constant
$\theta,\phi$:

@@

\begin{equation}\label{radial_string_length}
  \int_{r=r_1}^{r_2}\frac{dr}{\sqrt{1-2M/r}}=
  \left.
  \sqrt{r(r-2M)} +\frac{1}{2}\log\left(
  \frac{\sqrt{r}+\sqrt{r-2M}}{\sqrt{r}-\sqrt{r-2M}}\right)
  \right|_{r_1}^{r_2}.
  \end{equation}

Observe that integration may be performed with a lower limit
of~$r_1=2M$ with no difficulty, even though the radial component of
the Schwarzschild metric approaches infinity there.  Elementary
methods show that the proper length $\ell$ of radial strings has some
non-Euclidean features.  Firstly, consider the proper length from
$r=2M$ to $r=2M(1+w)$ where $w>0, \left|w\right|\ll 1$.  Then

@@

(compare the flat space string length of $2Mw$).  This would indicate
 that, close to the Schwarzschild radius $r=2M$, more string is
 required to connect two points with slightly differing radial
 coordinates than would be expected classically.  Taking a solar-mass
 black hole and $w=10^{-3}$ as an example, the proper length computed
 in flat space would be $10^{-2}\cdot 2M\simeq 3\,\mathrm{m}$; compare
 $\simeq 187\,\mathrm{m}$ from equation~\ref{ell}.  Passing to the
 case $w\gg 1$ we have an asymptotic expansion that begins

@@

There does not appear to be a simple analytical solution to
equation~\ref{rdashdash}; but it is a straightforward second order
ODE, amenable to numerical solution and no harder than the equivalent
for null geodesics, a standard homework problem.  Here we consider
only strings exterior to the event horizon, $r>2M$, which have a
well-defined point of closest approach to the singularity.  Because
the system is invariant under~$\phi\longrightarrow-\phi$, the string
must be symmetrical about this point.  If $r'$ is small, then a string
follows $r''=r-2M$, while tangential null geodesics follow $r''=r-3M$.
Working to first order, and writing~$d$ for the difference in
Schwarzschild~$r$ between string and null geodesic, we have $d''=M$.
Transforming coordinates to $(x,y) = (r\phi,r-r_0)$ and with boundary
conditions corresponding to both string and null geodesic being
tangential at the $(x,y)$ origin, we see that
$d=\frac{M}{2}(x/r_0)^2$.  However, a static local observer tied to
the string would measure
$d_\mathrm{local}=d/\sqrt{1-2M/r_0}$---applying
equation~\ref{schwarzschild}---and obtain
$d_\mathrm{local}=\frac{M}{2}(x/r_0)^2\left(1-2M/r_0\right)^{-1/2}$
which would correspond to a radius of curvature
of~$r_0\cdot\frac{r_0}{M}\cdot\left(1-2M/r_0\right)^{-1/2}$.  Thus a
taut string would {\em look} curved to a static observer in
Schwarzschild coordinates, in the sense that a tangential light beam
and taut string would diverge quadratically.  At the photon sphere
$r_0=3M$, such a string would appear to have a radius of~$\sqrt{27}M$.
For a solar mass black hole this would be about $15\unit{km}$, a
readily understandable result.
